$R^*$: A robust MCMC convergence diagnostic with uncertainty via gradient-boosted machines

### 2020-03-16 write_up/ms.tex
started discussion

Add another panel for the joint case

\begin{figure}[h]
	\centerline{\includegraphics[width=0.6\textwidth]{../output/unmixed_1.pdf}}
	\caption{\textbf{Chain prediction based on marginal distribution of a single parameter.} A shows the path of two chains that have mixed resulting in similar sampling distributions (to right of panel); B shows two chains that have not mixed.}
	\label{fig:marginal}
\end{figure}

\begin{figure}[h]
	\centerline{\includegraphics[width=0.4\textwidth]{../output/unmixed_2.pdf}}
	\caption{\textbf{Chain prediction based on joint distribution of two parameters.} Here, the marginal distributions of each dimensions are similar but the joint distributions differ.}
	\label{fig:joint}
\end{figure}

These two cases, whilst simple, illustrate the basis of our approach. To determine if a set of Markov chains have converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each sample belongs. By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$ where $N$ is the number of chains. By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\hat{R}$ \cite{gelman2013bayesian}. In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence. Algorithm \ref{alg:R_star} gives an explicit recipe for calculating $R^*$.

\begin{algorithm}[tb]
	\caption{$R^*$ calculation algorithm}
	\label{alg:R_star}
	\begin{algorithmic}
		\STATE Given $\theta^0$, $\epsilon$, $L$, $\cL, N$:
		\FOR{$m=1$ to $N$}
		\STATE Sample $k^0\sim\N(0, M)$.
		\STATE Set $\theta^m\leftarrow \theta^{m-1}, \tilde\theta\leftarrow \theta^{m-1}, \tilde k\leftarrow k^0$.
		\FOR{$i=1$ to $L$}
		\STATE Set $\tilde\theta, \tilde k \leftarrow \LF(\tilde\theta, \tilde k, \epsilon)$.
		\ENDFOR
		\STATE With probability $\alpha = \min\{1, \exp(H(\theta^{m-1}, k^0)-H(\theta^m,\tilde k))\}$
		set $\theta^m\leftarrow\tilde \theta$, $k^m\leftarrow -\tilde k$.
		\ENDFOR
		\STATE
		\STATE {\bf function} $\LF(\theta, k, \epsilon)$
		\STATE Set $\tilde k \leftarrow k +
		(\epsilon/2)\nabla_\theta\cL(\theta)$.
		\STATE Set $\tilde \theta \leftarrow \theta +
		\epsilon M^{-1}k'$.
		\STATE Set $\tilde k \leftarrow \tilde k +
		(\epsilon/2)\nabla_\theta\cL(\tilde\theta)$.
		\RETURN $\tilde \theta, \tilde k$.
	\end{algorithmic}
\end{algorithm}

The ML model we use here is a gradient-boosted regression tree (GBM), which was introduced in \cite{friedman2001greedy}, has been shown to be a highly predictive framework for use in tabular data \cite{chollet2018} like ours. Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s Caret package \cite{kuhn2008building}, which, in turn, uses the ``gbm'' package \cite{greenwell2019package}. The data for each chain has dimensions: $X\in \mathbb{R}^{SK}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters. We split each chain's samples into training and testing tests: here, we use 70\% of samples for training and 30\% for testing. The GBM model we use was found to be rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for models with 1000 samples) and its predictive performance was insensitive to its hyperparameters. In what follows, our hyperparameter settings were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations would be the minimum for each node and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).

@@

\begin{algorithm}[tb]
	\caption{Procedure to generate a single $R^*$ sample}
	\label{alg:R_star_uncertainty}
	\begin{algorithmic}
		\STATE Given $\theta^0$, $\epsilon$, $L$, $\cL, N$:
		\FOR{$m=1$ to $N$}
		\STATE Sample $k^0\sim\N(0, M)$.
		\STATE Set $\theta^m\leftarrow \theta^{m-1}, \tilde\theta\leftarrow \theta^{m-1}, \tilde k\leftarrow k^0$.
		\FOR{$i=1$ to $L$}
		\STATE Set $\tilde\theta, \tilde k \leftarrow \LF(\tilde\theta, \tilde k, \epsilon)$.
		\ENDFOR
		\STATE With probability $\alpha = \min\{1, \exp(H(\theta^{m-1}, k^0)-H(\theta^m,\tilde k))\}$
		set $\theta^m\leftarrow\tilde \theta$, $k^m\leftarrow -\tilde k$.
		\ENDFOR
		\STATE
		\STATE {\bf function} $\LF(\theta, k, \epsilon)$
		\STATE Set $\tilde k \leftarrow k +
		(\epsilon/2)\nabla_\theta\cL(\theta)$.
		\STATE Set $\tilde \theta \leftarrow \theta +
		\epsilon M^{-1}k'$.
		\STATE Set $\tilde k \leftarrow \tilde k +
		(\epsilon/2)\nabla_\theta\cL(\tilde\theta)$.
		\RETURN $\tilde \theta, \tilde k$.
	\end{algorithmic}
\end{algorithm}

@@

\subsection{Heterogeneity in chain variance}\label{sec:heterogeneity}
We generate four Markov chains, where each samples from an autoregressive order 1 (AR1) process of the form,
%
\begin{equation}
X_t = \rho X_{t-1} + \epsilon_t
\end{equation}
%
where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$. Three of the chains share the same $\sigma=\bar{\sigma}$, whereas the other chain has $\sigma=1/3\bar{\sigma}$, so that it has $1/3$ of the (unconditional) variance of the others. In each replicate, we simulate 1000 iterations from all four chains and fit a gradient-boosted model to a labelled training set of 2800 randomly-chosen samples, and use it to classify samples according to the chain which generated them in an independent test set comprising the remaining 1200 observations. We perform 1000 replicates and, in each case, we report the ratio of accuracy of the predictions to the null model which assigns observations to chains uniformly at random (i.e. having an accuracy of 25\%).

@@

\subsection{Poor convergence in joint distribution}
Simple AR1 example followed by 250 MVT normal.

@@

\subsection{Hierarchical model: Eight schools}
In this example, we examine a simple classic example used to highlight difficulties in performing sampling for hierarchical models: referred to as the "Eight schools" model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.

@@

\section{Discussion}
Add in index variable into set of predictors.
Relative insensitivity of GBM to settings (in all examples we used the same hyperparameters)

>>>>

\begin{figure}[h]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_1.pdf}}
	\caption{\textbf{Chain prediction based on the marginal distribution of a single parameter.} A shows the path of two chains that have mixed resulting in similar sampling distributions (to the right of panel); B shows two chains that have not mixed.}
	\label{fig:marginal}
\end{figure}

\begin{figure}[h]
	\centerline{\includegraphics[width=1.0\textwidth]{../output/unmixed_2.pdf}}
	\caption{\textbf{Chain prediction based on the joint distribution of two parameters where each chain's marginals are the same.} A shows the path of two chains that have mixed resulting in similar sampling distributions (to the right and above each panel); B shows two chains that have not mixed.}
	\label{fig:joint}
\end{figure}

These two cases, whilst simple, illustrate the basis of our approach. To determine if a set of Markov chains have converged to the same distribution, we train a supervised machine learning (ML) model to classify the chain to which each sample belongs. By evaluating its performance on an independent test set, we delineate whether chains have mixed based on whether classification accuracy is above the ``null'' case, where accuracy is $1/{N}$, and $N$ is the number of chains. By taking the ratio of ML accuracy to this null accuracy, we obtain a statistic that is interpretable in a similar way to $\hat{R}$ \cite{gelman2013bayesian}. In a nod to this established statistic, we call our statistic $R^*$, and, by design, $R^*\approx 1$ signifies convergence. Algorithm \ref{alg:R_star} gives an explicit recipe for calculating $R^*$.

\begin{algorithm}[tb]
	\caption{$R^*$ calculation}
	\label{alg:R_star}
	\begin{algorithmic}
		\STATE Given $\{X^{\{1\}},X^{\{2\}},...,X^{\{N\}}\}, S_\text{test}$:
		\FOR{$m=1$ to $N$}
		\STATE Partition, $X^{\{m\}}\rightarrow\{X^{\{m\}}_\text{train},X^{\{m\}}_\text{test}\}$
		\ENDFOR
		\STATE $X_\text{train} = \{X^{\{1\}}_\text{train},X^{\{2\}}_\text{train},...,X^{\{N\}}_\text{train}\}$
		\STATE Train ML model to classify chain id from any sample, $x$: $\text{ML}(x|X_\text{train}) \rightarrow c$
		\FOR{$s=1$ to $S_\text{test}$}
		\STATE Predict chain id, $c^\{s\} = \text{ML}(x^\{s\}|X_\text{train})$
		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
		\ENDFOR
		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
		\STATE Calculate ratio to null model, $R^* = \bar{a} / (1 / N) = N/\bar{a}$
		\RETURN $R^*$
	\end{algorithmic}
\end{algorithm}

The ML model we use here is a gradient-boosted regression tree (GBM), which was introduced in \cite{friedman2001greedy}, has been shown to be a highly predictive framework for use in tabular data \cite{chollet2018} like ours. Specifically, we use the GBM implementation in \textbf{\textsf{R}}'s Caret package \cite{kuhn2008building}, which, in turn, uses the ``gbm'' package \cite{greenwell2019package}. The data for each chain has dimensions: $X\in \mathbb{R}^{SK}$, where $S$ is the number of samples taken (here assumed the same for each chain, but this is not a binding constraint) and $K$ is the number of parameters. We split each chain's samples into randomly sampled training and testing tests: here, we use 70\% of samples for training and 30\% for testing. The GBM model we use was found to be rapid to execute training then prediction on the testing set (taking $<1$ second on a desktop computer for models with 1000 samples) and its predictive performance was insensitive to its hyperparameters. In what follows, our hyperparameter settings were: an interaction depth of 3, a shrinkage parameter of 0.1, 10 observations would be the minimum for each node and that 50 trees would be grown (see \url{https://github.com/ben18785/ml-mcmc-convergence} to replicate this functionality).

@@

\begin{algorithm}[tb]
	\caption{Procedure to generate $I$ samples of $R^*$}
	\label{alg:R_star_uncertainty}
	\begin{algorithmic}
		\STATE Given test data $X_\text{test}$, number of chains $N$, number of iterations $I$, and fitted
		\STATE model, $\text{ML}(x^{\{s\}}|X_\text{train})\rightarrow(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$:
		\FOR{$i=1$ to $I$}
		\FOR{$s=1$ to $S_\text{test}$}
		\STATE Predict chain id probabilities, $(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})= \text{ML}(x^\{s\}|X_\text{train})$
		\STATE Draw a chain id, $c^{\{s\}} \sim \text{categorical}(p_1^{\{s\}},p_2^{\{s\}},...,p_N^{\{s\}})$
		\STATE Compare with actual id, $c^s$: $a^{\{s\}}=\mathbbm{1}(c^{\{s\}}=c^s)$
		\ENDFOR
		\STATE Calculate predictive accuracy, $\bar{a} = \frac{1}{S_\text{test}} \sum_{s=1}^{S_\text{test}} a^{\{s\}}$
		\STATE Calculate ratio to null model, $R^{*i} = \bar{a} / (1 / N) = N/\bar{a}$
		\ENDFOR
		\RETURN $(R^{*1},R^{*2},...,R^{*I})$
	\end{algorithmic}
\end{algorithm}

@@

\subsection{Heterogeneity in chain variance: autoregressive example}\label{sec:heterogeneity}
We generate four Markov chains, where each samples from an autoregressive order 1 (AR1) process of the form,
%
\begin{equation}
X_t = \rho X_{t-1} + \epsilon_t
\end{equation}
%
where $\epsilon_t\stackrel{i.i.d.}{\sim}\mathcal{N}(0, \sigma)$. Three of the chains share the same $\sigma=\bar{\sigma}$, whereas the other chain has $\sigma=1/3\bar{\sigma}$, so that it has $1/3$ of the (unconditional) variance of the others. In each replicate, we simulate 1000 iterations from all four chains and fit a gradient-boosted model to a labelled training set of 2800 randomly-chosen samples, and use it to classify samples according to the chain which generated them in an independent test set comprising the remaining 1200 observations. We perform 1000 replicates and, in each case, we report the ratio of accuracy of the predictions to the null model which assigns observations to chains uniformly at random (i.e. having an accuracy of 25\%).

@@

\subsection{Lack of convergence of the joint distribution: multivariate normal model}\label{sec:multivariate_normal}
Simple AR1 example followed by 250 MVT normal.

@@

\subsection{Hierarchical model: Eight schools model}
In this example, we examine a simple classic example used to highlight difficulties in performing sampling for hierarchical models: referred to as the "Eight schools" model (see Section 5.5 in \cite{gelman2013bayesian}), which aimed to determine the effects of coaching on SAT scores in eight schools.

@@

\section{Discussion}
If an MCMC sampler has converged on the target distribution, the chains must be well ``mixed'', that is, given a sample it should be impossible to discern which chain generated that particular observation. Based on this observation, used supervised machine learning (ML) models to quantify the information about the generative chain identity contained in the samples. By taking the ratio of ML model predictive accuracy obtained on an independent test set to the accuracy of a null model (which predicts a chain's identity uniformly at random), this defines our $R^*$ statistic. By extracting the chain identity probabilities from each prediction in the test set, this allowed us to generate an uncertainty distribution for $R^*$. Across a range of previously published examples, $R^*$ was shown to be predictive of whether or not chains had converged.

The predominant methods for diagnosing MCMC convergence rely heavily on looking for differences in the marginal distributions of each dimension between chains. $R^*$ naturally includes this information in building a model capable of predicting the chain which generated each sample. It also naturally includes information about the joint distribution across all dimensions of the target. Since both the marginal and joint distributions will be approximately equal across converged chains, any measure of convergence should account for both of these aspects. Indeed, in \S\ref{sec:multivariate_normal}, we show that more established measures may indicate converge whereas $R^*$ shows otherwise. This is not a slight on existing measures, more that we argue this illustrates the complementarity of $R^*$ to them.

Add in index variable into set of predictors.

Relative insensitivity of GBM to settings (in all examples we used the same hyperparameters)
